Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Remove comments from Methods section, rewrite first part of Results section to correspond to the newly formulated simulation objectives.

This method is insufficient on two counts: 1) it may be challenging to
the untrained eye, and 2) only severely pathological cases of
non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
Therefore, a quantitative, diagnostic evaluation of convergence would be
preferred. Yet, monitoring convergence of iterative imputation
algorithms diagnostically is challenging. Iterative imputation
algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In
MCMC methods, convergence is not from a scalar to a point but from one
distribution to another. The values generated by the algorithm (e.g.,
imputed values) will vary even after convergence. Therefore, the aim of
convergence diagnostics for MCMC methods is not to establish the point
at which convergence is reached, but to monitor signs of non-convergence
\citep{hoff09}. Several of such diagnostics exist for MCMC methods, but
it is not known whether these are appropriate for iterative imputation
algorithms.

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the iterative imputation algorithm implemented
in the popular \texttt{mice} package \citep{mice} in \texttt{R}
\citep{R}. The convergence properties of the MICE algorithm are
investigated through model-based simulation. The results of this
simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets.

\textbf{Include sub-questions!} How can non-convergence be identified
diagnostically? Are common MCMC non-convergence diagnostics appropriate
for MICE? And if so, which threshold should be used to diagnose
non-convergence? How many iterations are sufficient/needed to be able to
diagnose non-convergence? Are the default number of iterations
sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
\textbf{How severe is it when the algorithm has not converged? And what
are guidelines for practice? Can the parameter of interest, estimand
\(Q\), be correct when the algorithm is not (yet) converged, and vice
versa?} \emph{(Maybe add these too? What are the effects of continued
iterations on estimates, predictions and inferences? Do the answers
differ with varying missingness proportions? That is, we vary the nr of
iterations and the missingness proportion because we assume that the
algorithm has not reached convergence at \(t=1\), and performs worse
with moderate to high missingness (not so much with little or a LOT of
missingness).)}

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on
\(k\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
(\(j = 1, 2, \dots, k\)) may be either observed or missing. \textbf{The
number of units \(i\) with at least one missing data value can be
divided by the total number of units \(n\) to obtain the `missingness
proportion' \(p_{mis}\).} The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
\(M \times T\) times plausible values, where \(M\) is the number of
imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
will be referred to as an `imputation chain'. \textbf{Add: \(\theta\)s
are scalar summaries of interest in the iterative algorithm (e.g., chain
means; the average of the imputed values in each imputation chain). }

@@

The goal of imputation through e.g.~FCS/MICE is to pbtain a multivariate
distribution by iterating over the sequence of univariate imputations.
Traditionally, the inspection of the multivariate convergence is done by
visually inspecting the univariate chain means and variances over the
iterations, or by tracking the Rhat and autocorrelation. The downside to
these approaches is that they either focus on the univariate state
space, or primarily track the change over the iterations of a
multivariate outcome conform the scientific model of interest. Focusing
on the convergence of these outcome parameters may influence this
procedure in the sense that the model of evaluation favors the model of
interest. Ideally, one would like to evaluate a model-independent
parameter that summarizes the multivariate nature of the data. We
propose eigenvalue as such a parameter, because it\ldots.

@@

\textbf{In Figure \ref{fig:diagnostics}A, the chain means from Figure
\ref{fig:non-conv} are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often much lower.
Therefore, we compute \(AC\) manually, see panel D. The AC values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence in the MICE algorithm on several quantities of
scientific interest \(Q\). Subsequently we will evaluate how well
\(\widehat{R}\) and \(AC\) perform in identifying the effects of
non-convergence. And finally, we will formulate an informed advice on
the requirements to safely assume \emph{sufficient} convergence in
practice. That is, we assume that the algorithm is \emph{sufficiently}
converged when each estimate \(\bar{Q}\) is an unbiased and confidence
valid estimate of the corresponding estimand \(Q\).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness
  proportion \(p_{mis}\) and a low number of iterations \(T\) will
  result in biased, invalid estimates of the quantities of scientific
  interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify
  signs of non-convergence in those simulation conditions where the
  \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of
  the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose
  non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\),
  \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too
  stringent for iterative imputation applications. In an empirical
  study, where \(\widehat{R}\) was used to inform the required
  imputation chain length, it took as many as 50 iterations to overcome
  the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet,
  scientific estimates were insensitive to continued iteration after
  \(t>5\) \citep{lace07}. We therefore suspect that \(\widehat{R}\) may
  over-estimate signs of non-convergence in iterative imputation
  algorithms. In contrast to this, it may occur that the signs of
  non-convergence are under-estimated by \(\widehat{R}\), in exceptional
  cases where the initial values of the algorithm are not appropriately
  over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
  values are chosen randomly from the observed data, hence we cannot be
  certain of over-dispersion in the initial values. In practice, we do
  not expect this to cause problems for identifying non-convergence with
  \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative
  imputation algorithms with typical convergence. That is, after only a
  few iterations, the randomness induced by the algorithm will
  effectively mitigate the risk of dependency within chains.
\end{enumerate}

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes across simulation runs 
\end{verbatim}

@@

In this study, sampling variance is not of interest. Therefore, a single
complete dataset may serve as comparative truth in all simulation
repetitions \citep{vink14}. The data-generating mechanism is a
multivariate normal distribution, representing person-data on three
predictor variables in a multiple linear regression problem
\textbf{(from an unspecified social scientific field of study)}. Let the
predictor space be defined as \[
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
\] A finite population of \(N=1000\) is simulated using the
\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
variable is constructed as outcome \(Y\). For each unit
\(i = 1, 2,..., N\), let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set,
we obtain the true values of each quantity of scientific interest \(Q\).

@@

We consider four types of \(Q\)s that are often of interest in empirical
research. Namely, two descriptive statistics: the mean \(\mu_j\) and
standard deviation \(\sigma_j\) of each variable
\(j = Y, X_1, X_2, X_3\). We also consider the regression coefficients
of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in
regression equation
\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\] where \(Y'\)
is the expected value of the outcome. And finally, the proportion of
variance explained by the set of predictors: coefficient of
determination \(r^2 \times 100\) \textbf{(note that lower case \(r\) is
used to avoid confusion with non-convergence diagnostic
\(\widehat{R}\))}. The true values of these scientific estimands are
calculated with the R functions \texttt{base::mean()},
\texttt{stats::sd()}, and \texttt{stats::lm()} \citep{R}. Each of these
\(Q\)s is estimated with a corresponding \(\bar{Q}\), after the complete
dataset is first \emph{amputed} (with a varying missingness proportion
\(p_{mis}\)) and subsequently imputed (with a varying number of
iterations \(T\)).

@@

The complete dataset is amputed with the \texttt{mice} package
\citep[function \texttt{mice::ampute()};][]{mice}. All missingness is
multivariate, and conform a `missing completely at random' missingness
mechanism \citep{rubin87}. I.e., the probability to be missing is the
same for all \(N \times k\) cells in \(y\), conditional on the
missingness proportion (\(p_{mis} =.05,.25,.5,.75,.95\)).

Missing datapoints in \(y\) are imputed with the \texttt{mice()}
function \citep{mice}. All imputation procedures are performed with
Bayesian linear regression imputation, and five imputation chains
(\(M=5\)). The number of iterations varies between simulation conditions
(\(T=1,2,...,100\)).

After each run of the \texttt{mice} algorithm, the function
\texttt{mice::complete()} is used to combine the imputed data
\(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five
completed datasets \{\(y_{obs}, y_{imp, m}\)\}, where
\(m = 1, 2, ..., 5\).

\begin{itemize}
\item
  on the 5 imputations \(y_{imp,m}\): get mean and var within imp chains
  as theta to compute conv diag --\textgreater{} then complete with
  \texttt{mice::complete()}
\item
  on the 5 completed datasets \{\(y_{obs}, y_{imp, m}\)\}: 1) estimate
  Qhats for mu and sigma --\textgreater{} pool as mean()
  --\textgreater{} get Qbars of mu and sigma; 2) compute lambda as theta
  for conv diag; 3) perform lm to get Qhats for betas and r\^{}2
  --\textgreater{} use Qhats of beta as theta to compute conv diag
  --\textgreater{} pool Qhats of beta and r\^{}2 to get Qbars
\end{itemize}

The estimator for each \(Q\) is \(\bar{Q}\)---the pooled aggregate of
the \(\hat{Q}\)s across imputations. To calculate the \(\bar{Q}\)s, we
combine the observed data \(y_{obs}\) and the imputed data for each
imputation \(y_{imp,m}\) with the function \texttt{mice::complete()}.
Descriptive statistics are computed as the average across imputations
for each variable (\(\mu_j\) and \(\sigma_j\), where
\(j = Y, X_1, X_2, X_3\)). Estimated regression coefficients are
obtained with the function \texttt{stats::lm()} for each imputation, and
then pooled conform \citet{vink14}. The coefficient of determination is
estimated for each imputation, and pooled using
\texttt{mice::pool.r.squared()}.

For each run of the imputation algorithm, we compute non-convergence
diagnostics \(\widehat{R}\) and \(AC\). We calculate \(\widehat{R}\) by
implementing Vehtari et al.'s \citeyearpar{veht19} recommendations, and
\(AC\) as the correlation between the \(t^{th}\) and the \((t+1)^{th}\)
iteration. The two methods are applied to four scalar summaries
\(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)),
chain variances (i.e., the variance in each \(y_{imp, m, j}\)), a
statistic of scientific interest \(\hat{Q}\) in each
\{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
\(\lambda_1\).

The \(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s
are evaluated against the bias in univariate mean estimates. Similarly,
the performance measure for \(\widehat{R}\) and \(AC\) applied to chain
variances is the bias in estimated standard deviations. To evaluate the
performance of \(\widehat{R}\) and \(AC\) on the eigenvalues, we will
use bias in the coefficient of determination, \(R^2\). Additionally, we
will evaluate the estimated regression coefficients, the coverage rate
of the CI95\% of the regression estimates, and the CI length
(\textbf{see definitions in old Methods}). \textbf{remove the term
performance measure here and add an extra paragraph to define
performance measures as bias in all estimands, and coverage rate and CI
length of regression coefficients.}

@@

The impact of inducing non-convergence in the iterative imputation
algorithm is assessed by comparing \(\bar{Q}\)s with \(Q\)s. For each
\(Q\), we compute bias as \(\bar{Q} - Q\). For the regression
coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage
rate (CR). CR is defined as the percentage of simulation repetitions in
which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the
true estimand \(Q\). Let
\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\] where
\(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. Since under-estimating the variance of
\(\bar{Q}\) may yield spurious inferences, we also inspect CI width
(CIW): the difference between the lower and upper bound of the 95\%
confidence interval around \(\bar{Q}\).

The performance of two methods is evaluated against the bias and
confidence validity of the estimated \(Q\)s.

@@

We want to know what the effect is of inducing non-conv of the bias in
all Qs and the confidence validity of the betas. So we plot the bias for
each Q against the number of iterations and the missingness proportion
in the condition. We look at averages across simulation repetitions. Is
there more bias in conditions where non-conv was induced? (NOPE for
univariates, YES for multivar.).

Then, we want to know if conditions in which the Qs are affected by
non-convergence (on average) are also identified by the conv diag. We do
this by looking at the bias in Q against the conv diag applied to
related thetas. Is the point at which there is no more bias the same
point at which the conv diag indicate minimal signs of non-conv? (NOPE,
they are too conservative).

Is one of the thetas better? (NAH, doesn't really matter).

What should the thresholds be? (LOWER than recommended) And the default
nr of iterarions? (MAYBE increase to 10?)

\textbf{Add to this section: 1) emphasize that the plots are averages
across repetitions, not within MICE; 2) sample effects due to single
complete dataset. 3) Add more info about figure legends and axes.)}

For reasons of brevity, we only discuss the convergence diagnostics for
the \(Q\) with the worst performance in terms of bias. For the
descriptve statistics, the magnitude of the bias was the largest in
\(Y\) (i.e, \(j = Y\) in \(\mu_j\) and \(\sigma_j\)). For the
\(Q=\beta_{1,2,3}\), \(\beta_1\) showed the worst perfomance (i.e, the
effect of \(X_1\) on \(Y\). Moreover, we only show conditions where
\(T\leq50\) in the figures. Full results are available from
\textbf{{[}link github repo{]}}.

>>>>

This method is insufficient on two counts: 1) it may be challenging to
the untrained eye, and 2) only severely pathological cases of
non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
Therefore, a quantitative, diagnostic evaluation of convergence would be
preferred---yet not straightforward. Iterative imputation algorithms
such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC
methods, convergence is not from a scalar to a point, but from one
distribution to another. The values generated by the algorithm (e.g.,
imputed values) will vary even after convergence. Since MCMC algorithms
do not reach a unique point at which convergence is established,
diagnostics we may only identify signs of non-convergence
\citep{hoff09}. Several of such diagnostics exist, but it is not known
whether these are appropriate for iterative imputation algorithms.

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the iterative imputation algorithm implemented
in the popular \texttt{mice} package \citep{mice} in \texttt{R}
\citep{R}. The convergence properties of the MICE algorithm are
investigated through model-based simulation. The results of this
simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets. \textbf{add societal relevance: what
is the consequence of non-convergence?}

\textbf{Include sub-questions and add structure of the paper!} How can
non-convergence be identified diagnostically? Are common MCMC
non-convergence diagnostics appropriate for MICE? And if so, which
threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{How severe is it when the algorithm
has not converged? And what are guidelines for practice? Can the
parameter of interest, estimand \(Q\), be correct when the algorithm is
not (yet) converged, and vice versa?} \emph{(Maybe add these too? What
are the effects of continued iterations on estimates, predictions and
inferences? Do the answers differ with varying missingness proportions?
That is, we vary the nr of iterations and the missingness proportion
because we assume that the algorithm has not reached convergence at
\(t=1\), and performs worse with moderate to high missingness (not so
much with little or a LOT of missingness).)}

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on
\(k\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\)
(\(j = 1, 2, \dots, k\)) may be either observed or missing. \textbf{The
number of units \(i\) in dataset \(y\) with at least one missing data
value, divided by the total number of units \(n\), is called the
missingness proportion in \(p_{mis}\).} The collection of observed data
values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is
referred to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample
\(M \times T\) times plausible values, where \(M\) is the number of
imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of
iterations (\(t = 1, 2, \dots, T\)). The collection of samples between
the initial value (at \(t=1\)) and the final imputed value (at \(t=T\))
will be referred to as an `imputation chain'. \textbf{Add: \(\theta\)s
are scalar summaries of interest in the iterative algorithm (e.g., chain
means; the average of the imputed values in each imputation chain). }

@@

\textbf{The goal of imputation through e.g.~MICE is to obtain a
multivariate distribution by iterating over the sequence of univariate
imputations. Traditionally, the inspection of the multivariate
convergence is done by visually inspecting the univariate chain means
and variances over the iterations, or by tracking the \(\widehat{R}\)
and autocorrelation. The downside to these approaches is that they
either focus on the univariate state space, or primarily track the
change over the iterations of a multivariate outcome conform the
scientific model of interest. Focusing on the convergence of these
outcome parameters may influence this procedure in the sense that the
model of evaluation favors the model of interest. Ideally, one would
like to evaluate a model-independent parameter that summarizes the
multivariate nature of the data. We propose eigenvalue as such a
parameter, because it\ldots. }

@@

\textbf{In Figure \ref{fig:diagnostics}A, the chain means from Figure
\ref{fig:non-conv} are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often much lower.
Therefore, we compute \(AC\) manually, see panel D. The \(AC\) values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence in the MICE algorithm on several quantities of
scientific interest \(Q\). Subsequently we will evaluate how well
\(\widehat{R}\) and \(AC\) perform in identifying the effects of
non-convergence (\textbf{add thetas}). And finally, we will formulate an
informed advice on the requirements to safely assume \emph{sufficient}
convergence in practice. That is, we assume that the algorithm is
\emph{sufficiently} converged when each estimate \(\bar{Q}\) is an
unbiased and confidence valid estimate of the corresponding estimand
\(Q\). (\textbf{add to formulate advice on thetas?})

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness
  proportion \(p_{mis}\) and a low number of iterations \(T\) will
  result in biased, invalid estimates of the quantities of scientific
  interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify
  signs of non-convergence in those simulation conditions where the
  \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of
  the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose
  non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\),
  \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too
  stringent for iterative imputation applications. In an empirical
  study, where \(\widehat{R}\) was used to inform the required
  imputation chain length, it took as many as 50 iterations to overcome
  the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet,
  scientific estimates were insensitive to continued iteration after
  \(t>5\) \citep{lace07}. We therefore suspect that \(\widehat{R}\) may
  over-estimate signs of non-convergence in iterative imputation
  algorithms, compared to the bias the estimates. In contrast to this,
  it may occur that signs of non-convergence are under-estimated by
  \(\widehat{R}\), in exceptional cases where the initial values of the
  algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}.
  In e.g.~\texttt{mice}, initial values are chosen randomly from the
  observed data, hence we cannot be certain of over-dispersion in the
  initial values. In practice, we do not expect this to cause problems
  for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative
  imputation algorithms with typical convergence. That is, after only a
  few iterations, the randomness induced by the algorithm should
  effectively mitigate the risk of dependency within chains.
\item
  \textbf{Something about thetas}
\end{enumerate}

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes of all simulation runs 
\end{verbatim}

@@

In this study, sampling variance is not of interest. Therefore, a single
complete dataset may serve as comparative truth in all simulation
repetitions \citep{vink14}. The data-generating mechanism is a
multivariate normal distribution, representing person-data on three
predictor variables in a multiple linear regression problem
\textbf{(from an unspecified social scientific field of study)}. Let the
predictor space be defined as \[
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
\] A finite population of \(N=1000\) is simulated using the
\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
variable is constructed as outcome \(Y\). For each unit
\(i = 1, 2,..., N\), let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set,
we obtain the true values of several quantities of scientific interest,
\(Q\)s. \textbf{MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?}
\(Q\)s are the targets that will be estimated in each simulation
repetition, after first \emph{amputing} the complete data with varying
missingness proportions, then imputing the missingness with a varying
number of iterations, then estimating the \(Q\)s on the \(M\) completed
datasets, and finally pooling the \(M\) \(\hat{Q}\)s to obtain a
\(\bar{Q}\) for each \(Q\). \textbf{Add that we try to predict the
effect of non-converegence (i.e., the difference between \(Q\) and
\(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and
\(AC\), by appliying these methods to several scalar summaries of the
state space of the algorithm, \(\theta\)s.}

@@

We consider four types of \(Q\)s that are often of interest in empirical
research. Namely, two descriptive statistics: the mean \(\mu_j\) and
standard deviation \(\sigma_j\) of each variable
\(j = Y, X_1, X_2, X_3\). We also consider the regression coefficients
of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in
regression equation
\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\] where \(Y'\)
is the expected value of the outcome. And finally, the proportion of
variance explained by the set of predictors: coefficient of
determination \(r^2 \times 100\) \textbf{(note that lower case \(r\) is
used to avoid confusion with non-convergence diagnostic
\(\widehat{R}\))}. The true values of these scientific estimands are
calculated on the complete data with the R functions
\texttt{base::mean()}, \texttt{stats::sd()}, and \texttt{stats::lm()}
\citep{R}.

@@

In each simulation repetition, we ampute the complete dataset five times
to obtain five different incomplete datasets, \(y\)s. The ratio between
\(y_{obs}\) and \(y_{mis}\) in each \(y\) depends on the missingness
proportion: \(p_{mis} =.05,.25,.5,.75,.95\). We ampute the complete data
with the \texttt{mice} package \citep[function
\texttt{mice::ampute()};][]{mice}. All missingness is multivariate, and
conform a `missing completely at random' missingness mechanism
\citep{rubin87}. I.e., the probability to be missing is the same for all
\(N \times k\) cells in \(y\), conditional on the missingness proportion
\(p_{mis}\).

Missing datapoints in the incomplete datasets, \(y\)s, are imputed with
the \texttt{mice()} function \citep{mice}. All imputation procedures are
performed with Bayesian linear regression imputation, and five
imputation chains (\(M=5\)). Because \(M=5\), each run of the
\texttt{mice()} function results in five sets of imputations,
\(y_{imp,m}\)s, where \(m = 1, 2, ..., 5\). The number of iterations in
each run of the algorithm varies between simulation conditions
(\(T=1,2,...,100\)).

From the 5 imputations \(y_{imp,m}\): get chain means and chain
variances from imputation object to use as scalar summary \(\theta\) in
convergence diagnostics. Then use \texttt{mice::complete()} to combine
the imputed data \(y_{imp,m}\) with the observed data \(y_{obs}\),
resulting in five completed datasets \{\(y_{obs}, y_{imp, m}\)\}.

On the 5 completed datasets \{\(y_{obs}, y_{imp, m}\)\}, we:

\begin{itemize}
\item
  Estimate \(\hat{Q}\)s for \(Q= \mu\) and \(Q=\sigma\) with
  \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. Then pool
  the \(\hat{Q}\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
\item
  Compute \(\lambda_1\), to use as \(\theta\) in convergence
  diagnostics. By definition, the first eigenvalue of the
  variance-covariance matrix, \(\lambda_1\), is equal to the variance of
  the principal component solution of each \{\(y_{obs}, y_{imp, m}\)\},
  which we calculate with \texttt{stats::princomp()} \citep{R}.
\item
  Perform multiple linear regression with \texttt{stats::lm()} \citep{R}
  to get the \(\hat{Q}\)s for \(Q=\beta\) and \(Q=r^2\). Pool the
  \(\hat{Q}\)s conform \citet{vink14} to get \(\bar{Q}\)s for the
  \(\beta\)s and \(r^2\) (\textbf{check pooling of \(r^2\), instead of
  \texttt{mice::pool.r.squared()}}). Additionally, use the \(\hat{Q}\)s
  of \(Q=\beta\) as \(\theta\) to compute convergence diagnostics.
\end{itemize}

We apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) to each
scalar summary \(\theta\). We calculate \(\widehat{R}\) by implementing
Vehtari et al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the
correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration.

@@

We assess the impact of inducing non-convergence in the iterative
imputation algorithm by comparing \(\bar{Q}\)s with \(Q\)s. For each
\(Q\), we calculate bias as \(\bar{Q} - Q\). For the regression
coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage
rate (CR). CR is defined as the percentage of simulation repetitions in
which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the
true estimand \(Q\). Let
\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\] where
\(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. If we obtain nominal coverage (CR = 95\%), we
can conclude that the \(\bar{Q}\)s are conficence valid estimates of the
\(Q\)s. Finally, we inspect CI width (CIW): the difference between the
lower and upper bound of the 95\% confidence interval around
\(\bar{Q}\). CIW is of interest because too short CIs indicate
under-estimation of the variance in \(\bar{Q}\), which may yield
spurious inferences.

With these performance measures, we also evaluate how well the
non-convergence diagnostics \(\widehat{R}\) and \(AC\) can identify
simulation conditions in which the \(\bar{Q}\)s are \emph{not} unbiased,
confidence valid estimates of the \(Q\)s. \textbf{Rephrase and add
corresponding \(Q\)s:} The two methods are applied to four scalar
summaries \(\theta\): chain means (i.e., the mean in each
\(y_{imp, m, j}\)), chain variances (i.e., the variance in each
\(y_{imp, m, j}\)), the estimated regression coefficients
\(\beta_{1,2,3}\) in each \{\(y_{obs}, y_{imp, m}\)\} (i.e., a
user-defined statistic of scientific interest \(\hat{Q}\)), and the
first eigenvalue of the variance-covariance matrix in each
\{\(y_{obs}, y_{imp, m}\)\}, \(\lambda_1\).

@@

\textbf{Add to this section: 1) emphasize that the plots are averages
across repetitions, not within MICE; 2) sample effects due to single
complete dataset?? 3) Add more info about figure legends and axes.)}

Primarily, we want to know what the effect is of inducing
non-convergence on the estimated \(Q\)s. Specifically, we are interested
in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of
\(\bar{Q}\) for \(Q=\beta\). Therefore, we plot these performance
measures against the simulation conditions (the number of iterations
\(T\) and the missingness proportion \(p_{mis}\)). We look at averages
across simulation repetitions.

Secondly, we want to know if conditions in which the \(Q\)s are affected
by non-convergence are also identified by the non-convergance
diagnostics \(\widehat{R}\) and \(AC\). We do this, again, by looking at
the performance measures for the \(Q\)s, but now we evaluate these
against what \(\widehat{R}\) and \(AC\) indicate about signs of non-the
convergence in the \(\theta\)s.

\begin{itemize}
\item
  Is there more bias in conditions where non-convergence was induced?
  (NOPE for univariates, YES for multivar.).
\item
  Is the point of \emph{sufficient} convergence (at which the
  \(\bar{Q}\)s are unbiased and confidence valid estimates of the
  \(Q\)s) the same point as identified by the non-convergence
  diagnostics? (NOPE, they are too conservative).**
\item
  Is one of the thetas better? (NAH, doesn't really matter).
\item
  What should the thresholds be? (LOWER than recommended).
\item
  And the default nr of iterarions? (MAYBE increase to 10?).
\end{itemize}
